---
title: "Look at your objects"
date: "2034-08-05"
draft: true
---

Random thought today: There are a lot of ways to "check in" on your intermediate objects in R.

![](behold my stuff)

It's definitely good practice and something I have trouble pushing my students to do.  Maybe I need to be more deliberate about how to do it.


So, there's the classic way of just printing it out.  This is fine.  I tend to peek at my objects this way, except I do the peeking in the console... I can NOT get my students to adopt a workflow that pops between notebook and console though.  Maybe it's not the best.

```{r}
bob <- 1:10
bob
```

---

I also see this in some folks' code:

```{r}
bob <- 1:10; bob
```

I especially see it in plotting with ggplot for some reason:

```{r}
library(tidyverse)
library(palmerpenguins)

p <- ggplot(penguins, aes(x = body_mass_g)) +
  geom_histogram(); p
```

I do not like this at all.  Perhaps it's a bias against semicolons, I thought I left those behind when I gave up on Java.  But I don't like the print statement being hidden on a line with code.

---

Now, a student taught me this trick, and I think it's super rad:

```{r}
(bob <- 1:10)
```

But it does get a bit inelegant/cumbersome with multiline code and pipelines in my opinion:

```{r}
(pen_ad <- penguins %>%
  filter(species == "Adelie") %>%
   summarize(mean(body_mass_g, na.rm = TRUE)))
```

---

Speaking of pipelines, I'm on the fence about the best way to "check in" on progress of a long pipeline.  I tend to just highlight part of the pipeline and Cmd+Enter to run that section.  But that's kinda unreproducible and also gets annoying if I'm doing it many times.

Students tend to delete or comment out segments of pipelines and I do NOT like this, it's so unwieldy.

---

`magrittr` has a cute pipe `%T>%` that means "do this next step but don't pass its results", which we can use in conjunction with `print()` to check stuff.

It's almost perfect but the necessity of `print()` and the 

```{r}
library(magrittr)

penguins %>%
  filter(species == "Adelie") %T>%
  print() %>%  
  summarize(n_rows = n())
         
```

(Honestly, I wish we in the tidyverse sphere used the other `magrittr` pipes more.  Maybe another mini-post one day...)

---

One option is to throw a `peek()` inside a pipeline...
